A comparison of different methods for calculating tangent-stiffness matrices in a massively parallel computational peridynamics code.

### 2014-03-05 jacobian_calc_CMAME.tex
spellcheck

While, for simplicity, the examples shown in the following refer only to linear
problems, tangent-stiffness matrices generally arise in the context of
non-linear analysis where Newton's method or quasi-Newton methods are used
in the minimization of a given residual function.  In this context, the
tangent-stiffness matrix can be thought of as the linearization of the system
about a particular configuration.  It has been shown
\cite{hughes1978consistent,hughes1978unconditionally} that in order to preserve
the quadratic convergence properties of the Newton's methods, it is necessary
that this linearization is carried out in a manner that is \emph{consistent}
with the algorithmic constraint equations used when computing the internal
forces that arise due to deformations.  An example of these constraint
equations would be the Kuhn-Tucker conditions \cite{simo1998} that are used in
integration of a flow rule for plasticity modeling.  If the continuum
tangent-moduli are used in place of the \emph{consistent} or \emph{algorithmic
tangent moduli} in the solution of a non-linear solid mechanics problem,
convergence may still be achieved, but not in the quadratic manner that makes
Newton's method attractive for this class of problems.  In the setting of
non-linear solid mechanics, there is a very small class of material model
algorithms in which consistent tangent-stiffness operators can be derived
analytically.  Perfect plasticity and plasticity with isotropic hardening used
in combination with a general nearest-point projection or \emph{radial return
algorithm} are a few examples.  Therefore, when general models (and certainly
more complex ones) are implemented into a general purpose computational
mechanics code, the tangent-stiffness operators are typically defined via a
numerical approximation.

@@

\subsubsection{Implementing the complex-step method in Peridigm} 
\label{subsubsec:Impl}
%
The CS method was implemented in \emph{Peridigm} in a manner that follows
closely the implementation of the FD technique.  However, because of
\emph{Peridigm}'s reliance on \emph{Trilinos} and specifically the
\emph{Epetra} package, special steps had to be followed to allow the use of
complex number data types.  This is because \emph{Epetra} vectors, which can be
thought of as distributed memory parallel cousins to standard C++
\change[Michael]{STL}{Standard Template Library} vectors, are hard coded to be
of type {\tt double} and can not be simply declared as having a complex data
type. \note[Michael]{re: I.3}\add[Michael]{For the reader's convenience,
    \emph{Epetra} vectors and STL vectors are what are known as 'container
    classes', whose purpose in object oriented computer programming is to
    automate the creation, accessing, modification and destruction of the
    underlying contained data. Templated container classes allow for user
    choice in what properties that underlying data exhibits.  For example, in
    this work, complex arithmetic cannot be performed on the data contained
    within an \emph{Epetra} vector because that data, of type {\tt double} does
    not have the property of following the rules of complex arithmetic. The
    reason the non-templated \emph{Epetra} vector could not be dispensed with
    for the purposes of this work is because the \emph{Epetra} vector was
tightly integrated with \emph{Peridigm's} parallel communication strategy and
linear algebra code.} Program methods that were used in the course of applying
the FD method were copied, renamed and re-written to follow the CS formula.
Because of restrictions imposed by the container class being used, this
involved changing the data type of intermediate variables which were locally
scoped, and dynamically casting persistent memory variables to complex data
types where necessary. \note[Michael]{re: I.3}\add[Michael]{For the reader's
convenience, 'locally scoped' in the C++ programming language refers to
data which is created within a context more specific than the context of
the entire program, or 'translation unit'.  Most data created this way is
destroyed when that specific context is left. The topic of contexts in C++
is very broad, so for the purposes of this work context refers to a program
method or function that begins, concludes and yields a result, also
destroying locally scoped data. Dynamic casting for the purposes of this
work refers to copying the value of data contained within an \emph{Epetra}
vector into locally scoped data that has the property of following the rules of
complex arithmetic, where a mapping called 'type casting' allows the
translation of data between different types.} No attempt was made to ``tune''
the CS code in any way for performance. The algorithm  was implemented in such
as way that it was in one-to-one correspondence with the FD and CD methods;
however, it must be noted that the overhead associated with the dynamic casts
in the CS method likely hurts its overall performance and this issue could be
alleviated by using the next generation of templated \emph{TPetra} vectors
available in \emph{Trilinos} that are capable of taking an explicit
instantiation of any data type including {\tt complex}. \note[Michael]{re:
I.3}\add[Michael]{For the purposes of this work, an explicit instantiation
refers to requesting the creation of an object, or instance, of a templated
class or container class that holds data which has the property of
following desired mathematic or logic rules.} The version of \emph{Peridigm}
modified to implement CS can be found in the paper repository.

@@

The methods were compared by running two sets of test problems where each
method would solve a problem concurrently. In this context, concurrently means
successively, drawing from the same independent variables, yet within the same
computational process; this definition is characterized in further detail in
Section~\ref{JGAM}. Each test problem simulated a $\SI{4}{\meter}$ by
$\SI{0.5}{\meter}$ by $\SI{0.5}{\meter}$ meter block of a material undergoing
tension along the axis parallel to the long dimension of the block. The test
problems were set up in \emph{Peridigm} as quasistatic equilibrium problems,
where the displacement boundary conditions used to apply tension were applied
gradually in ``load steps'', and an equilibrium solution for each load step was
achieved before applying the next load step. Applying the load
in gradual steps is a widely used technique to ensure the Newton's method is
always initialized near the actual solution and therefore likely to converge.
The fictional material represented in the model had values of
$\SI{1.515e4}{\mega\pascal}$ for bulk modulus, and $\SI{7.813e4}{\mega\pascal}$
for shear modulus. The peridynamic horizon, a length scale that effectively
sets the bandwidth of tangent-stiffness matrix in the computation, used in the test
problems was always set to three times the nominal discretization size (i.e.
node spacing in the particle discretization scheme). This results in a minimum
tangent-stiffness matrix bandwidth of 21 for a 3D problem (for an optimal node
numbering scheme). Pictured in Fig.~\ref{fig:SCMesh} is a viewgraph for one of
the peridynamic meshes used in this study. The coloration in the image
indicates node numbering, where a smooth color transition means that the nodes
have close ID numbers. The abrupt color changes in the center and at the ends
of the body show that numbering is not contiguous and is assigned with respect to four regioins. The purpose of designating
these regions, for this study, was to apply displacement boundary
conditions called \emph{volume constraints} in peridynamics.

@@

\section{Results and discussion}
\label{sec:Res}
%
Table~\ref{tab:results} shows the compute core count, number of nonzero tangent-stiffness (TS) matrix elements, load step average calculation time and load step average accuracy. The accuracy measure is given by Eqn.~(\ref{eqn:accuracy}) and the choice of elastic moduli mentioned in Section~\ref{tcs}.
%
\begin{table*}[!ht]    
  \scriptsize
  \centering
        \caption{Averaged Results, Each Test} \label{tab:results}   
       \begin{tabular}{c c c c c c c c c}
         \toprule
         Cores & TS Elements & \multicolumn{4}{c}{Calculation Time ($\si{\second}$)} & \multicolumn{3}{c}{Accuracy Difference ($\si{\mega\pascal}$)} \\ 
         & $\times 10^6$ & CS & CD & FD & AD & CS & CD & FD \\
        \midrule
        1 & $\num{1.99}$  & 3.5 & 2.7 & 1.7 & 1.6 & $\num{1.92E-10}$ & $\num{1.21E-4}$ & .137 \\
        1 & $\num{4.25}$  & 6.2& 4.9& 3.2& 2.9 & $\num{2.28E-10}$ & $\num{9.94E-4}$ & .148 \\
        1 & $\num{7.79}$  & 11.2& 8.9& 5.7& 5.2 & $\num{2.38E-10}$ & $\num{1.59E-4}$ & .145\\
        1 & $\num{15.4}$  & 26.7& 21.0& 13.4& 12.2 & $\num{2.33E-10}$ & $\num{4.61E-4}$ & .12 \\
        1 & $\num{20.0}$  & 28.1& 22.2& 14.4& 13.1 & $\num{2.76E-10}$ & $\num{1.05E-3}$ & .145 \\
        1 & $\num{31.4}$  & 47.6& 37.7& 24.2& 21.9 & $\num{2.64E-10}$ & $\num{1.65E-3}$ & .133 \\
        1 & $\num{40.1}$  & 55.6& 44.1& 28.4& 25.9 & $\num{3.03E-10}$ & $\num{1.92E-3}$ & .148 \\
        1 & $\num{83.9}$  & 138.9& 109.6& 70.2& 64.0 & $\num{3.63E-10}$ & $\num{1.64E-3}$ & .123 \\
        1 & $\num{165.0}$  & 277.3& 218.1& 139.4& 126.5 & $\num{3.26E-10}$ & $\num{2.18E-3}$ & .128 \\
        32 & $\num{1670}$  & 336.1& 277.7& 200.6& 233.1 & $\num{6.21E-10}$ & $\num{1.52E-2}$ & .176 \\
        64 & $\num{1670}$  & 169.9& 140.7& 102.0& 119.7 & $\num{6.20E-10}$ & $\num{1.50E-2}$ & .177 \\
        96 & $\num{1670}$  & 114.7& 95.0& 69.1 & 79.7 & $\num{6.18E-10}$ & $\num{1.50E-2}$ & .177 \\
        128 & $\num{1670}$  & 86.4& 71.8& 52.4 &58.8 & $\num{6.16E-10}$ & $\num{1.47E-2}$ & .177 \\
        \bottomrule
    \end{tabular}
\end{table*}

@@

Additionally, there are Quasi-Newton methods such as Broyden's method which
uses an approximate inverse Jacobian and updates it with information from the 
iterate and residual \cite{dennis1971convergence}. A method called
Incomplete Jacobian Newton's method may the closest to the idea of deliberately
making use of an inaccurate Jacobian in the sense that arbitrary elements of
the full Jacobian are set to zero to reduce computational effort
\cite{liu2008incomplete}. The Quasi-Newton methods represent a compromise of
Newton's method for the promise of faster iterations in exchange for a lower
convergence rate. For both of these methods, the iterate and residual are
computed as in Newton's method. As to the question of algorithmic consistency,
these numerical methods evaluate the same linearized constitutive law and
governing equation as for the exact Newton's method which in the case of a
plastic simulation are involved in iteratively determining algorithmic tangent moduli. 
In fact, a secant equation based method is appropriate for Newton's method with algorithmic
tangent moduli since the Jacobian ought only depend on known converged
values such that the Newton iterate could not artificially drive the plasticity
of the model \cite[Chap. 6]{belytschko1999nonlinear}. It should be noted that
constraints or a line search algorithm to minimize the residual are necessary
whereas the pure secant method does not return unique solutions in
multi-dimensions.
%

>>>>

While, for simplicity, the examples shown in the following refer only to linear
problems, tangent-stiffness matrices generally arise in the context of
non-linear analysis where Newton's method or quasi-Newton methods are used
in the minimization of a given residual function.  In this context, the
tangent-stiffness matrix can be thought of as the linearization of the system
about a particular configuration.  It has been shown
\cite{hughes1978consistent,hughes1978unconditionally} that in order to preserve
the quadratic convergence properties of the Newton's methods, it is necessary
that this linearization is carried out in a manner that is \emph{consistent}
with the algorithmic constraint equations used when computing the internal
forces that arise due to deformations.  An example of these constraint
equations would be the Kuhn-Tucker conditions \cite{simo1998} that are used in
integration of a flow rule for plasticity modeling.  If the continuum
tangent-modulii are used in place of the \emph{consistent} or \emph{algorithmic
tangent moduli} in the solution of a non-linear solid mechanics problem,
convergence may still be achieved, but not in the quadratic manner that makes
Newton's method attractive for this class of problems.  In the setting of
non-linear solid mechanics, there is a very small class of material model
algorithms in which consistent tangent-stiffness operators can be derived
analytically.  Perfect plasticity and plasticity with isotropic hardening used
in combination with a general nearest-point projection or \emph{radial return
algorithm} are a few examples.  Therefore, when general models (and certainly
more complex ones) are implemented into a general purpose computational
mechanics code, the tangent-stiffness operators are typically defined via a
numerical approximation.

@@

\subsubsection{Implementing the complex-step method in \emph{Peridigm}} 
\label{subsubsec:Impl}
%
The CS method was implemented in \emph{Peridigm} in a manner that follows
closely the implementation of the FD technique.  However, because of
\emph{Peridigm}'s reliance on \emph{Trilinos} and specifically the
\emph{Epetra} package, special steps had to be followed to allow the use of
complex number data types.  This is because \emph{Epetra} vectors, which can be
thought of as distributed memory parallel cousins to standard C++
\change[Michael]{STL}{Standard Template Library} vectors, are hard coded to be
of type {\tt double} and can not be simply declared as having a complex data
type. \note[Michael]{re: I.3}\add[Michael]{For the reader's convenience,
    \emph{Epetra} vectors and STL vectors are what are known as 'container
    classes', whose purpose in object oriented computer programming is to
    automate the creation, accessing, modification and destruction of the
    underlying contained data. Templated container classes allow for user
    choice in what properties that underlying data exhibits.  For example, in
    this work, complex arithmetic cannot be performed on the data contained
    within an \emph{Epetra} vector because that data, of type {\tt double} does
    not have the property of following the rules of complex arithmetic. The
    reason the non-templated \emph{Epetra} vector could not be dispensed with
    for the purposes of this work is because the \emph{Epetra} vector was
tightly integrated with \emph{Peridigm's} parallel communication strategy and
linear algebra code.} Program methods that were used in the course of applying
the FD method were copied, renamed and re-written to follow the CS formula.
Because of restrictions imposed by the container class being used, this
involved changing the data type of intermediate variables which were locally
scoped, and dynamically casting persistent memory variables to complex data
types where necessary. \note[Michael]{re: I.3}\add[Michael]{For the reader's
convenience, 'locally scoped' in the C++ programming language refers to
data which is created within a context more specific than the context of
the entire program, or 'translation unit'.  Most data created this way is
destroyed when that specific context is left. The topic of contexts in C++
is very broad, so for the purposes of this work context refers to a program
method or function that begins, concludes and yields a result, also
destroying locally scoped data. Dynamic casting for the purposes of this
work refers to copying the value of data contained within an \emph{Epetra}
vector into locally scoped data that has the property of following the rules of
complex arithmetic, where a mapping called 'type casting' allows the
translation of data between different types.} No attempt was made to ``tune''
the CS code in any way for performance. The algorithm  was implemented in such
as way that it was in one-to-one correspondence with the FD and CD methods;
however, it must be noted that the overhead associated with the dynamic casts
in the CS method likely hurts its overall performance and this issue could be
alleviated by using the next generation of templated \emph{TPetra} vectors
available in \emph{Trilinos} that are capable of taking an explicit
instantiation of any data type including {\tt complex}. \note[Michael]{re:
I.3}\add[Michael]{For the purposes of this work, an explicit instantiation
refers to requesting the creation of an object, or instance, of a templated
class or container class that holds data which has the property of
following desired mathematic or logic rules.} The version of \emph{Peridigm}
modified to implement CS can be found in the paper repository.

@@

The methods were compared by running two sets of test problems where each
method would solve a problem concurrently. In this context, concurrently means
successively, drawing from the same independent variables, yet within the same
computational process; this definition is characterized in further detail in
Section~\ref{JGAM}. Each test problem simulated a $\SI{4}{\meter}$ by
$\SI{0.5}{\meter}$ by $\SI{0.5}{\meter}$ meter block of a material undergoing
tension along the axis parallel to the long dimension of the block. The test
problems were set up in \emph{Peridigm} as quasi-static equilibrium problems,
where the displacement boundary conditions used to apply tension were applied
gradually in ``load steps'', and an equilibrium solution for each load step was
achieved before applying the next load step. Applying the load
in gradual steps is a widely used technique to ensure the Newton's method is
always initialized near the actual solution and therefore likely to converge.
The fictional material represented in the model had values of
$\SI{1.515e4}{\mega\pascal}$ for bulk modulus, and $\SI{7.813e4}{\mega\pascal}$
for shear modulus. The peridynamic horizon, a length scale that effectively
sets the bandwidth of tangent-stiffness matrix in the computation, used in the test
problems was always set to three times the nominal discretization size (i.e.
node spacing in the particle discretization scheme). This results in a minimum
tangent-stiffness matrix bandwidth of 21 for a 3D problem (for an optimal node
numbering scheme). Pictured in Fig.~\ref{fig:SCMesh} is a viewgraph for one of
the peridynamic meshes used in this study. The coloration in the image
indicates node numbering, where a smooth color transition means that the nodes
have close ID numbers. The abrupt color changes in the center and at the ends
of the body show that numbering is not contiguous and is assigned with respect to four regions. The purpose of designating
these regions, for this study, was to apply displacement boundary
conditions called \emph{volume constraints} in peridynamics.

@@

\section{Results and discussion}
\label{sec:Res}
%
Table~\ref{tab:results} shows the compute core count, number of nonzero tangent-stiffness (TS) matrix elements, load step average calculation time and load step average accuracy. The accuracy measure is given by Eqn.~(\ref{eqn:accuracy}) and the choice of elastic modulii mentioned in Section~\ref{tcs}.
%
\begin{table*}[!ht]    
  \scriptsize
  \centering
        \caption{Averaged Results, Each Test} \label{tab:results}   
       \begin{tabular}{c c c c c c c c c}
         \toprule
         Cores & TS Elements & \multicolumn{4}{c}{Calculation Time ($\si{\second}$)} & \multicolumn{3}{c}{Accuracy Difference ($\si{\mega\pascal}$)} \\ 
         & $\times 10^6$ & CS & CD & FD & AD & CS & CD & FD \\
        \midrule
        1 & $\num{1.99}$  & 3.5 & 2.7 & 1.7 & 1.6 & $\num{1.92E-10}$ & $\num{1.21E-4}$ & .137 \\
        1 & $\num{4.25}$  & 6.2& 4.9& 3.2& 2.9 & $\num{2.28E-10}$ & $\num{9.94E-4}$ & .148 \\
        1 & $\num{7.79}$  & 11.2& 8.9& 5.7& 5.2 & $\num{2.38E-10}$ & $\num{1.59E-4}$ & .145\\
        1 & $\num{15.4}$  & 26.7& 21.0& 13.4& 12.2 & $\num{2.33E-10}$ & $\num{4.61E-4}$ & .12 \\
        1 & $\num{20.0}$  & 28.1& 22.2& 14.4& 13.1 & $\num{2.76E-10}$ & $\num{1.05E-3}$ & .145 \\
        1 & $\num{31.4}$  & 47.6& 37.7& 24.2& 21.9 & $\num{2.64E-10}$ & $\num{1.65E-3}$ & .133 \\
        1 & $\num{40.1}$  & 55.6& 44.1& 28.4& 25.9 & $\num{3.03E-10}$ & $\num{1.92E-3}$ & .148 \\
        1 & $\num{83.9}$  & 138.9& 109.6& 70.2& 64.0 & $\num{3.63E-10}$ & $\num{1.64E-3}$ & .123 \\
        1 & $\num{165.0}$  & 277.3& 218.1& 139.4& 126.5 & $\num{3.26E-10}$ & $\num{2.18E-3}$ & .128 \\
        32 & $\num{1670}$  & 336.1& 277.7& 200.6& 233.1 & $\num{6.21E-10}$ & $\num{1.52E-2}$ & .176 \\
        64 & $\num{1670}$  & 169.9& 140.7& 102.0& 119.7 & $\num{6.20E-10}$ & $\num{1.50E-2}$ & .177 \\
        96 & $\num{1670}$  & 114.7& 95.0& 69.1 & 79.7 & $\num{6.18E-10}$ & $\num{1.50E-2}$ & .177 \\
        128 & $\num{1670}$  & 86.4& 71.8& 52.4 &58.8 & $\num{6.16E-10}$ & $\num{1.47E-2}$ & .177 \\
        \bottomrule
    \end{tabular}
\end{table*}

@@

Additionally, there are Quasi-Newton methods such as Broyden's method which
uses an approximate inverse Jacobian and updates it with information from the 
iterate and residual \cite{dennis1971convergence}. A method called
Incomplete Jacobian Newton's method may the closest to the idea of deliberately
making use of an inaccurate Jacobian in the sense that arbitrary elements of
the full Jacobian are set to zero to reduce computational effort
\cite{liu2008incomplete}. The Quasi-Newton methods represent a compromise of
Newton's method for the promise of faster iterations in exchange for a lower
convergence rate. For both of these methods, the iterate and residual are
computed as in Newton's method. As to the question of algorithmic consistency,
these numerical methods evaluate the same linearized constitutive law and
governing equation as for the exact Newton's method which in the case of a
plastic simulation are involved in iteratively determining algorithmic tangent moduli. 
In fact, a secant equation based method is appropriate for Newton's method with algorithmic
tangent modulii since the Jacobian ought only depend on known converged
values such that the Newton iterate could not artificially drive the plasticity
of the model \cite[Chap. 6]{belytschko1999nonlinear}. It should be noted that
constraints or a line search algorithm to minimize the residual are necessary
whereas the pure secant method does not return unique solutions in
multi-dimensions.
%
